**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Update code and text for The PacBio error profile section

regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
clrs <- rainbow(length(regions))
pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
names(clrs) <- regions
```

@@

```{r error_profile}
sub.matrix <- read.table(file=paste("pipeline_dev/", regions[1], "/", regions[1], ".mock.filter.error.matrix", sep=""), header=T, row.names=1)[-7,]
for(i in 2:length(regions)){
	sub.matrix <- sub.matrix + read.table(file=paste("pipeline_dev/", regions[i], "/", regions[i], ".mock.filter.error.matrix", sep=""), header=T, row.names=1)[-7,]
}

matches <- c(sub.matrix[1,1], sub.matrix[2,2], sub.matrix[3,3], sub.matrix[4,4])
total <- sum(sub.matrix)
errors <- sub.matrix
diag(errors[1:4,1:4]) <- 0
init.error.rate <- 100 * sum(errors) / total				#1.925717

subst.bias <- apply(errors[1:4,1:4], 2, sum)/apply(sub.matrix[1:4,1:4], 2, sum)
subst.bias <- 100 * subst.bias/sum(subst.bias)
#       rA        rT        rG        rC
# 22.43918 27.21620 24.06557 26.27904
#these are more or less equal to each other

insert.bias <- (sub.matrix[,"rGap"]/apply(sub.matrix, 1, sum))[1:4]
insert.bias <- 100 * insert.bias/sum(insert.bias)
#       qA       qT       qG       qC
# 23.46112 24.36667 24.33064 27.84157
#these are more or less equal to each other

delete.bias <- (sub.matrix["qGap",]/apply(sub.matrix, 2, sum))[1:4]
delete.bias <- 100 * delete.bias/sum(delete.bias)
#           rA      rT       rG       rC
#qGap 11.43234 9.46907 44.57506 34.52353
#interestingly the G's and C's are more likely to be deleted than the As or Ts

substitutions <- 100*sum(errors[1:4,1:4])/sum(errors)	#35.80084
insertions <- 100*sum(errors[,5])/sum(errors)			#45.27368
deletions <- 100*sum(errors[5,])/sum(errors)			#17.27259
ambiguous <- 100*sum(errors[6,])/sum(errors)			#2.101505

# Need to connect error types with the quality scores
error.quality <- read.table(file=paste("pipeline_dev/", regions[1], "/", regions[1], ".mock.filter.error.quality", sep=""), header=T, row.names=1)[-7,]
for(i in 2:length(regions)){
	error.quality <- error.quality + read.table(file=paste("pipeline_dev/", regions[i], "/", regions[i], ".mock.filter.error.quality", sep=""), header=T, row.names=1)[-7,]
}
error.quality["72",] <- error.quality["72",] + error.quality["80",]
error.quality <- error.quality[1:72,]

#get quantiles
matches <- rep(as.numeric(rownames(error.quality)), error.quality$matches)
m.quant <- quantile(matches, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))

subs <- rep(as.numeric(rownames(error.quality)), error.quality$substitutions)
s.quant <- quantile(subs, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))

ins <- rep(as.numeric(rownames(error.quality)), error.quality$insertions)
i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))

ambig <- rep(as.numeric(rownames(error.quality)), error.quality$ambiguous)
a.quant <- quantile(ambig, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
```
***The PacBio error profile.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(init.error.rate, digits=3, nsmall=2)`%. Insertions, deletions, substitutions, and ambiguous base calls
accounted for `r format(insertions, digits=3, nsmall=1)`,
`r format(deletions, digits=3, nsmall=1)`, `r format(substitutions, digits=3, nsmall=1)`,
and `r format(ambiguous, digits=2, nsmall=1)`% of
the errors, respectively. The substitution errors were equally likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(delete.bias["rG"], digits=3, nsmall=1)`%) and cytosines
(`r format(delete.bias["rC"], digits=3, nsmall=1)`%) were more likely to be deleted
than adenines (`r format(delete.bias["rA"], digits=3, nsmall=1)`%) or thymidines
(`r format(delete.bias["rT"], digits=2, nsmall=1)`%). When we considered the Phred
quality score of each base call, we observed a median quality score of `r m.quant["50%"]`
for correct base calls and scores of `r s.quant["50%"]` and `r i.quant["50%"]` for
substitutions and insertions, respectively (Figure 1A). Although there
was a broad distribution of quality scores with each type of base call, the errors could
largely be distinguished from the correct base calls.

@@

***A basic sequence curation procedure.***
To establish a simple curation procedure, we culled any sequence that contained
an ambiguous base call, had a string of the same base repeated 9 or more times,
did not start and end at the expected alignment coordinates for that region
of the 16S rRNA gene, or that was chimeric. This reduced the experiment-wide
error rate from `r format(init.error.rate, digits=3, nsmall=2)` to
`r format(composite.basic.error, digits=2, nsmall=2)`%. This basic procedure
resulted in the removal of between
`r format(round(min(100*(1-reasons.lost[,"good"])), digits=1), nsmall=1)`
(`r pretty.region[names(which.min(100*(1-reasons.lost[,"good"])))]`) and
`r round(max(100*(1-reasons.lost[,"good"])), digits=1)`
(`r pretty.region[names(which.max(100*(1-reasons.lost[,"good"])))]`)% of the
reads. The percentage of reads removed increased with the length of the
fragment (Figure 2). The number of reads removed
because of the presence of ambiguous base calls was similar to the number of
reads that were removed for not fully aligning to the correct region within the
16S rRNA gene (Table 2). The latter class of errors was
generally due to sequence truncations that could not be explained.

@@

**Figure 1. Summary of errors in data generated using PacBio sequencing platform
to sequence various regions within the 16S rRNA gene.** Quality scores varied
with error types (A). The sequencing error rate of the amplified gene fragments
increased with mismatches to the barcodes and primers (B). The sequencing error
rate declined with increasd sequencing coverage; however, increasing the
sequencing depth beyond 10-fold coverage had no meaningful effect on the
sequencing error rate (C).

>>>>

regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
clrs <- rainbow(length(regions))
pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
names(clrs) <- regions

mock_error <- read.table(file="data/process/mock.error.report")
mock_error_nochim <- mock_error[mock_error$numparents==1,]
mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
```

@@

```{r error_profile}
mutations <- c("AA", "AT", "AG", "AC", "TA", "TT", "TG", "TC", "GA", "GT", "GG", "GC", "CA", "CT", "CG", "CC", "Ai", "Ti", "Gi", "Ci", "dA", "dT", "dG", "dC")

mutation_table <- mock_error_good[, which(colnames(mock_error_good)%in%mutations)]
mutation_count <- apply(mutation_table, 2, sum)

mutation_total <- sum(mutation_count)
matches <- sum(mutation_count[c("AA", "TT", "GG", "CC")])
mis_matches <- mutation_total - matches
init_error_rate <- 100 * mis_matches / mutation_total		#0.6514488

#substitutions
substitutions <- mutations[grep("[ATGC].", mutations)]
substitutions <- substitutions[grep(".[ATGC]", substitutions)]
substitutions <- substitutions[! substitutions %in% c("AA", "TT", "GG", "CC")]
substitution_counts <- mutation_count[substitutions]
substitution_total <- sum(substitution_counts)
substitution_bias <- substitution_counts / substitution_total

reference <- gsub(".(.)", "\\1", names(substitution_bias))
# aggregate(substitution_bias, by=list(reference), mean)
#
# Group.1          x
# 1       A 0.10192188
# 2       C 0.05944286
# 3       G 0.07434043
# 4       T 0.09762816

substitution_rate <- 100 * substitution_total / mis_matches

# [1] 50.90346

#deletions
deletion_counts <- mutation_count[grep("d.", mutations)]
deletion_total <- sum(deletion_counts)
deletion_bias <- 100 * deletion_counts / deletion_total

#dA        dT        dG        dC
#24.25770 17.99132 39.44415 18.30683

deletion_rate <- 100 * deletion_total / mis_matches

#[1] 17.93361

#insertions
insertion_counts <- mutation_count[grep(".i", mutations)]
insertion_total <- sum(insertion_counts)
insertion_bias <- 100 * insertion_counts / insertion_total

#Ai        Ti        Gi        Ci
#23.13509 19.63951 30.05906 27.16633

insertion_rate <- 100 * insertion_total / mis_matches

#[1] 31.16293

# Need to connect error types with the quality scores
error_quality <- read.table(file="data/process/mock.quality.report", header=T, row.names=1)

#get quantiles
matches <- rep(as.numeric(rownames(error_quality)), error_quality$matches)
m.quant <- quantile(matches, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
m.94 <- 100 * sum(matches >= 94)/length(matches)
# [1] 80.52649

subs <- rep(as.numeric(rownames(error_quality)), error_quality$substitutions)
s.quant <- quantile(subs, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
s.94 <- 100 * sum(subs >= 94)/length(subs)
# [1] 79.96451

ins <- rep(as.numeric(rownames(error_quality)), error_quality$insertions)
i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
i.94 <- 100 * sum(ins >= 94)/length(ins)
# [1] 80.35283
```

***The PacBio error profile.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(100*init_error_rate, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
accounted for `r format(insertion_rate, digits=3, nsmall=1)`,
`r format(deletion_rate, digits=3, nsmall=1)`, and `r format(substitution_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(deletion_bias["dG"], digits=3, nsmall=1)`%) and adenines
(`r format(deletion_bias["dA"], digits=3, nsmall=1)`%) were more likely to be deleted
than cytosines (`r format(deletion_bias["dC"], digits=3, nsmall=1)`%) or thymidines
(`r format(deletion_bias["dT"], digits=2, nsmall=1)`%). The PacBio quality values varied between
`r paste(range(c(matches, subs, ins)), collapse=' and ')`. Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(m.94, digits=3, nsmall=1)`%), substitutions (`r format(s.94, digits=3, nsmall=1)`%), and insertions (`r format(i.94, digits=3, nsmall=1)`%). Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(cor(mock_error_good$error, mock_error_good$pred_error), digits=2, nsmall=2)`; Figure 1A).

@@

***A basic sequence curation procedure.***
To establish a simple curation procedure, we culled any sequence that contained
an ambiguous base call, had a string of the same base repeated 9 or more times,
did not start and end at the expected alignment coordinates for that region
of the 16S rRNA gene, or that was chimeric. This reduced the experiment-wide
error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
`r format(composite.basic.error, digits=2, nsmall=2)`%. This basic procedure
resulted in the removal of between
`r format(round(min(100*(1-reasons.lost[,"good"])), digits=1), nsmall=1)`
(`r pretty.region[names(which.min(100*(1-reasons.lost[,"good"])))]`) and
`r round(max(100*(1-reasons.lost[,"good"])), digits=1)`
(`r pretty.region[names(which.max(100*(1-reasons.lost[,"good"])))]`)% of the
reads. The percentage of reads removed increased with the length of the
fragment (Figure 2). The number of reads removed
because of the presence of ambiguous base calls was similar to the number of
reads that were removed for not fully aligning to the correct region within the
16S rRNA gene (Table 2). The latter class of errors was
generally due to sequence truncations that could not be explained.

@@

[revise - Fig 1A plot predicted error rate vs observed error rate]
**Figure 1. Summary of errors in data generated using PacBio sequencing platform
to sequence various regions within the 16S rRNA gene.** Quality scores varied
with error types (A). The sequencing error rate of the amplified gene fragments
increased with mismatches to the barcodes and primers (B). The sequencing error
rate declined with increasd sequencing coverage; however, increasing the
sequencing depth beyond 10-fold coverage had no meaningful effect on the
sequencing error rate (C).
